\sysfull

### 2018-04-15 paper/adaptiveModel.tex
add brain dump about Coala the execution model

\textcolor{red}{THIS BRAIN DUMP ABOUT THE EXECUTION MODEL - Placeholder}

>>>>

\subsection {Motivating the Design of \sys's Execution Model.} The challenge of \emph{efficiently} executing a program under frequent power failures, lies in the fact that the program needs to be energy aware. This challenge becomes even bigger when \emph{generality} is a design requirement, and consequently, the system must not require a specific hardware circuitry to offer its efficiency. \emph{Flexibility} is a key for wide adaptation, therefore, \sys's execution model should not limit the programmer to a subset of the functionality of the underlying software technology. \emph{\sys champions an efficient, general, and flexible intermittent execution.}

### 2018-04-15 paper/background.tex
limitations of usability by Alpaca added

%\begin{figure}
\begin{wrapfigure}{t!}{0.5\textwidth}
	\begin{subfigure}[t]{\linewidth}
		\centering \includegraphics[width=\columnwidth]{figures/crc_example}
		\caption{Simplified C code snippet of a CRC calculation from~\cite{hicks_mibench2_2016}: per-byte message division by a polynomial; \texttt{NV} denotes non-volatile variable declaration.}\label{fig:crc_example}
	\end{subfigure}
	\begin{subfigure}[t]{\linewidth}
		\centering \includegraphics[width=\columnwidth]{figures/crc_example_war}
		\caption{Execution steps of the loop body in the snippet above: non-volatile checkpointing did not guarantee data consistency as data has been manipulated (line 9) with stale reminder (line 3)}\label{fig:crc_example_war}
	\end{subfigure}
	\caption{\textbf{Code example demonstrating effect of write after read on volatile memory checkpointing.}}\label{fig:code_demo_incosistency}
\end{wrapfigure}
%\end{figure}

@@

Energy harvesting devices operate using energy extracted from sources such as radio frequency transmissions and solar energy. These devices elide tethered power or a battery, instead collecting energy into a capacitor, operating when sufficient energy accumulates, and upon depleting the energy, turning off and recharging.
\TODO{We need a figure that shows what intermittent execution looks like.}
%
%Batteryless operation has a number of important advantages, making intermittent computing an important research domain. Supplying power to billions~\cite{gartner_iot} of embedded computers using batteries is not sustainable. The European Commission estimates that more than 160 kilotons of consumer batteries enter the European Union annually~\cite{eu_batteries_2016}. Batteries are an environmental risk, are fragile, are limited in their number of charge/discharge cycles, and may require costly physical maintenance that is difficult or impossible deployed (e.g. in space~\cite{kicksat}). By contrast, super-capacitors are durable, promising millions of charge/discharge cycles~\cite[Sec. I]{ongaro_pwre_2012}. The main limitation in moving to capacitor-based energy storage is that capacitor energy density---and consequently operating discharge time---is orders of magnitude less than a battery. 
%
%Given current technology development, battery-less systems are best suited for very long-term sensing and monitoring where access to recharge is either prohibitive or impossible. These include battery-less image capture and processing~\cite{naderiparizi_rfid_2015}, animal monitoring~\cite{thomas_jbcs_2012} or implantable~\cite{rodriguez_tbcs_2015} and digestible~\cite{nadeau_naturebio_2017} sensors.
%
Many platforms enable intermittent, battery-less energy harvesting-based computation. For instance, computational RFIDs---open-source TI MSP430-based~\cite{wolverine} WISP~\cite{wisp5} (with its variants such as WISPCam~\cite{naderiparizi_rfid_2015}, NFC-WISP~\cite{zhao_rfid_2015} or NeuralWISP~\cite{holleman_biocas_2008}), Moo~\cite{moo}, and commercial ones such as~\cite{medusa_farsens_2017}. Other intermittently-powered platforms include ambient backscatter tag~\cite{liu_sigcomm_2013,parks_sigcomm_2014} or battery-less phone~\cite{talla_imwut_2017}.

@@

\textcolor{red}{reference work by Intel~\cite{baghsorkhi_cgo_2018}}

%Energy-harvesting devices execute software according to the {\em intermittent execution model}. Physically, a device charges until a threshold, operates briefly until its energy is depleted, shuts down recharges, and repeats the cycle.
%
Software running on an energy-harvesting device executes {\em intermittently} because buffered energy is only available \emph{sometimes}. An intermittent execution is composed of operating periods interspersed with power failures~\cite{dino,chain,alpaca,ratchet}. The frequency of failures depends on the size of the device's energy storage buffer: a larger buffer allows longer operating periods. Energy-harvesters provide input power orders of magnitude less than operating power, making recharging negligible during operation.

Intermittent execution is different from continuous execution. A power failure clears volatile  state (registers, stack, and globals) and non-volatile memory (e.g., FRAM) persists. At a failure, control flows to a prior point in the execution: by default, to the beginning of {\tt main()}. Early intermittent systems preserved progress by periodically checkpointing volatile execution context to non-volatile memory~\cite{mementos,quickrecall}, sometimes using hardware support~\cite{mementos,mottola2017harvos,hibernusplusplus,hibernus,idetic}.

@@

Checkpoints of volatile state preserve intermittent progress, but do not ensure data consistency~\cite{dino,chain,ratchet}. Data may become inconsistent if an attempt to execute some computation {\em writes} to a non-volatile variable, then power fails, then a second attempt to re-execute the same computation incorrectly {\em reads} the value written in the first attempt, rather than the variable's original value. The situation occurs when code includes a WAR dependence between operations that manipulate non-volatile variables~\cite{ratchet,dino,alpaca}.

Figure~\ref{fig:code_demo_incosistency} illustrates how state can become inconsistent in an intermittent execution using the cyclic redundancy check (CRC) code from MIBench2~\cite{hicks_mibench2_2016}. The code computes the CRC for an $\texttt{nBytes}$ byte message $\texttt{msg}$ with remainder $\texttt{Rmnd}$.

Despite checkpointing at line 3 in Fig.~\ref{fig:crc_example}, the code may compute  $\texttt{data}$ incorrectly because of the read of $\texttt{Rmnd}$ on line 4 and the write of $\texttt{Rmnd}$ on line 5. Figure~\ref{fig:crc_example_war} shows an intermittent execution. The second iteration writes $\texttt{Rmnd}$, but power fails before the checkpoint. After restarting, line 9 reads the updated value of $\texttt{Rmnd}$, producing an incorrect $\texttt{data}$ value. Checkpoint-based systems risk violating memory consistency~\cite{dino}. To maintain consistency, prior systems~\cite{dino,ratchet} {\em version} a subset of non-volatile data with the checkpoint. Task-based systems~\cite{chain,alpaca}, which we focus on, ensure consistency using programmer, compiler, and runtime support.

@@

Task-based execution models~\cite{dino,chain,alpaca} ask the programmer to decompose their program into tasks. A task is a function with no caller containing arbitrary computation, sensing, and communication. A programmer describes task control-flow as a {\em task graph}. Task flow happens at programmer demarcated transition points that may be conditional on program values. Task-based programming abstractions guarantee that tasks execute {\em atomically}, regardless of power failures. Task-based runtime systems ensure task-atomic semantics by ensuring that repeated task executions are idempotent. The key to idempotence is ensuring that non-volatile updates made by an interrupted task are never visible to a future task execution.

@@

Despite its advantages, the task-based programming model for intermittently-powered computing devices has its limitations. Two major ones being: {\em memory overhead} and {\em inflexibility to changing energy conditions}.

@@

>>>>

%\begin{figure}
\begin{wrapfigure}{t!}{0.5\textwidth}
	\begin{subfigure}[t]{\linewidth}
		\centering \includegraphics[width=\columnwidth]{figures/crc_example}
		\caption{Simplified C code snippet of a CRC calculation from~\cite{hicks_mibench2_2016}: per-byte message division by a polynomial; \texttt{NV} denotes non-volatile variable declaration.}\label{fig:crc_example}
	\end{subfigure}
	\begin{subfigure}[t]{\linewidth}
		\centering \includegraphics[width=\columnwidth]{figures/crc_example_war}
		\caption{Execution steps of the loop body in the snippet above: non-volatile checkpointing did not guarantee data consistency as data has been manipulated (line 9) with stale reminder (line 3)}\label{fig:crc_example_war}
	\end{subfigure}
	\caption{\textbf{Code example demonstrating checkpointing gives rise to data inconsistency.}}\label{fig:code_demo_incosistency}
\end{wrapfigure}
%\end{figure}

@@

Energy harvesting devices operate using energy extracted from sources such as radio frequency transmissions and solar energy. These devices elide tethered power or a battery, instead collecting energy into a capacitor, operating when sufficient energy accumulates, and upon depleting the energy, turning off and recharging.
%\TODO{We need a figure that shows what intermittent execution looks like.}
%
%Batteryless operation has a number of important advantages, making intermittent computing an important research domain. Supplying power to billions~\cite{gartner_iot} of embedded computers using batteries is not sustainable. The European Commission estimates that more than 160 kilotons of consumer batteries enter the European Union annually~\cite{eu_batteries_2016}. Batteries are an environmental risk, are fragile, are limited in their number of charge/discharge cycles, and may require costly physical maintenance that is difficult or impossible deployed (e.g. in space~\cite{kicksat}). By contrast, super-capacitors are durable, promising millions of charge/discharge cycles~\cite[Sec. I]{ongaro_pwre_2012}. The main limitation in moving to capacitor-based energy storage is that capacitor energy density---and consequently operating discharge time---is orders of magnitude less than a battery. 
%
%Given current technology development, battery-less systems are best suited for very long-term sensing and monitoring where access to recharge is either prohibitive or impossible. These include battery-less image capture and processing~\cite{naderiparizi_rfid_2015}, animal monitoring~\cite{thomas_jbcs_2012} or implantable~\cite{rodriguez_tbcs_2015} and digestible~\cite{nadeau_naturebio_2017} sensors.
%
There are several energy harvesting battery-less hardware platforms. For instance, computational RFIDs---open-source TI MSP430-based~\cite{wolverine} WISP~\cite{wisp5} (with its variants such as WISPCam~\cite{naderiparizi_rfid_2015}, NFC-WISP~\cite{zhao_rfid_2015} or NeuralWISP~\cite{holleman_biocas_2008}), Moo~\cite{moo}, and commercial ones such as~\cite{medusa_farsens_2017}. Other platforms include ambient backscatter tag~\cite{liu_sigcomm_2013,parks_sigcomm_2014} or battery-less phone~\cite{talla_imwut_2017}.

@@

%\textcolor{red}{reference work by Intel~\cite{baghsorkhi_cgo_2018}}

%Energy-harvesting devices execute software according to the {\em intermittent execution model}. Physically, a device charges until a threshold, operates briefly until its energy is depleted, shuts down recharges, and repeats the cycle.
%
Software running on an energy-harvesting device executes {\em intermittently} because energy sources are not always available to harvest and buffer sufficient operational energy. An intermittent execution is composed of operating periods interspersed with power failures~\cite{dino,chain,alpaca,ratchet}. The frequency of failures depends on the size of the device's energy storage buffer: a larger buffer allows longer operating periods. 
%Energy-harvesters provide input power orders of magnitude less than operating power, making recharging negligible during operation.
%Intermittent execution is different from continuous execution. 
A power failure clears volatile  state (registers, stack, and globals) and non-volatile memory (e.g., FRAM) persists. Upon a power failure, control flows to a prior point in the execution: by default, to the beginning of {\tt main()}. Early intermittent systems preserved progress by periodically checkpointing volatile execution context to non-volatile memory~\cite{mementos,quickrecall}, sometimes using hardware support~\cite{mementos,mottola2017harvos,hibernusplusplus,hibernus,idetic}.

@@

Checkpoints of volatile state preserve intermittent progress, but do not ensure data consistency~\cite{dino,chain,ratchet}. Data may become inconsistent if an attempt to execute some computation {\em writes} to a non-volatile variable, then power fails, then a second attempt to re-execute the same computation incorrectly {\em reads} the value written in the first attempt, rather than the variable's original value. The situation occurs when code includes a WAR dependence between operations that manipulate non-volatile variables~\cite{ratchet,dino,alpaca}---see Figure~\ref{fig:code_demo_incosistency} that illustrates how state can become inconsistent in an intermittent execution using the cyclic redundancy check (CRC) code from MIBench2~\cite{hicks_mibench2_2016}.
%Figure~\ref{fig:code_demo_incosistency} illustrates how state can become inconsistent in an intermittent execution using the cyclic redundancy check (CRC) code from MIBench2~\cite{hicks_mibench2_2016}. The code computes the CRC for an $\texttt{nBytes}$ byte message $\texttt{msg}$ with remainder $\texttt{Rmnd}$.
%Despite checkpointing at line 3 in Fig.~\ref{fig:crc_example}, the code may compute  $\texttt{data}$ incorrectly because of the read of $\texttt{Rmnd}$ on line 4 and the write of $\texttt{Rmnd}$ on line 5. Figure~\ref{fig:crc_example_war} shows an intermittent execution. The second iteration writes $\texttt{Rmnd}$, but power fails before the checkpoint. After restarting, line 9 reads the updated value of $\texttt{Rmnd}$, producing an incorrect $\texttt{data}$ value. Checkpoint-based systems risk violating memory consistency~\cite{dino}. 
To maintain consistency, prior systems~\cite{dino,ratchet} {\em version} a subset of non-volatile data with the checkpoint. Task-based systems~\cite{chain,alpaca}, which we focus on, ensure consistency using programmer, compiler, and runtime support.

@@

Task-based execution models~\cite{dino,chain,alpaca} ask the programmer to decompose their program into tasks. A task is a function with no caller containing arbitrary computation, sensing, and communication. A programmer describes task control-flow as a {\em task graph}. Task flow happens at programmer demarcated transition points that may be conditional on program values. 
%Task-based programming abstractions guarantee that tasks execute {\em atomically}, regardless of power failures. 
Task-based runtime systems ensure task-atomic semantics by ensuring that repeated task executions are idempotent. The key to idempotence is ensuring that non-volatile updates made by an interrupted task are never visible to a future task execution.

@@

Despite their advantages, the task-based programming models have their limitations. Two major ones being: {\em memory overhead} and {\em inflexibility to dynamic energy harvesting conditions}.

@@

%Przemek placeholder
%\textbf{Task-based Models have Syntactic Limitations.} State of the art task-based programming models, Alpaca~\cite{alpaca}, has substantial limitations that prohibit writing generic programs. First, for Alpaca (and Chain~\cite{chain}), accesses to protected variables must be applied \emph{only} by using the variable's name, i.e. reference by variable address is not possible. This limitation implies that existing third-party libraries, such as MCU-tailored FFT library~\cite{}, cannot be implemented to support Alpaca or Chain (not to mention the impossibility of the development of an intermittent-safe \texttt{memcpy}). Furthermore, if a protected variable needs to be accessed in a code path that is not predictable by the compiler, for instance an Interrupt Service Routine, Alpaca cannot protect that access.
